How to publish in Biological Sciences: a guide for the uninitiated

### 2021-12-15 1.03-transparency.Rmd
Add files via upload
# with lots of changes following copy edits and proofs

There need for transparency in science stems from the fact that most societies are making the majority of their decisions based on evidence coming through the guidance of science. If scientific evidence is at the heart of decision making, then the collection of this evidence must be transparent to those who make resulting policies, together with those who challenge them in a democracy. Moreover, as public funded scientists we should be setting a community standard of transparency for the rest of society to follow. Publishing of science is currently in transition, and it is vital that the scientific community leads the way forwards, and that we are not led by for-profit publishers. One of the ways to achieve transparency is through [preregistration](#commitment) of your research project to avoid [confirmation bias](#PruittData), which is, in part, a product of commercial publishers and the [metrics they promote](#impactfactor). In order to make this effective, we need the [gatekeepers](#gatekeepers) of our journals to support the [preregistration](#commitment) of research hypotheses and methods. Right now, journals should be openly advocating and encouraging preregistration with a plan to transition their journal in future to a system embracing rigour, reproducibility, and transparency [RRT @valdez2020improving]. However, many editors are resisting this move as they feel that there is no support from the community. This may well be the case, but inequalities in science, and particularly in publishing, mean that editors can either be instruments of change or at the heart of inequality in publishing (see [Part IV](#editors)). Either our editors will lead us toward transparency, or we as a community simply need to demand that they change their practices. Currently, editors are responding to calls for transparency by making small steps [for example asking for open coding: @powers2019open], rather than adopting transparency wholesale through the badge system set up by Kidwell et al. [-@kidwell2016badges].

@@

### Accepting the alternative hypothesis
At the outset of our scientific research we pose a hypothesis with the expectation that we will be able to accept or reject our null hypothesis. We often think of rejecting the null hypothesis as the only result that we are interested in, but if we only ever reported these results we would not be responsible in moving our field forwards. That is, in a world where we only report significant results (i.e. reject the null hypothesis) we would necessarily keep repeating experiments where the null hypothesis is accepted, because there would never be the evidence that the hypothesis had been previously tested in the literature. This is called confirmation bias, and it’s actually practised by the majority of scientific journals who won’t consider a null result. It’s easy to see why this is a bad policy, but it is the prevailing culture in science.

If journals only publish manuscripts that reject the null hypothesis [_cf_ @franco2014publication], researchers are more likely to mine their data for positive results (P hacking), or re-write their hypothesis in order to reject the null (HARKing) [@measey2021how]. Deceptive practices such as p hacking, HARKing and [salami-slicing](#salami) are not in the interests of any journals, or the scientific project in general [@ioannidis2005why; @forstmeier2017detecting; @measey2021how].

@@

### Novel research
One criterion for many journals is that the research should be novel. This is increasingly practiced by journal editors as you move up the [Impact Factor](#impactfactor) levels. Novelty sells (just think of the meaning of “new” in newspaper), and that’s the basis for selling novel stories from higher Impact Factor journals. The perils of testing increasingly unlikely hypotheses and how this inflates Type II errors as well as increasing the proportion of Type I errors are widely acknowledged [@forstmeier2017detecting; @measey2021how]. Novelty also stifles repeatability. If we can never repeat studies in science, then a fundamental tenet of the methodology is repressed. Reproducibility in science has received a lot of attention recently, as attempts to reproduce the results of highly cited research have failed (cites). This has been followed by general outrage among scientists that things should change [@anderson2007normative; @munafo2017manifesto], including a majority of those in biological sciences [@baker20161500]. The irony that these reports and requests are published in exactly the journals that will refuse to publish research that seeks to repeat work (is not novel) is clearly lost on the editors. However, more nuanced views are also coming forwards to actively introduce variable conditions and sampling of biological variation into the study design to more fully represent the nature of biological variation making studies more likely to be replicated [@voelkl2020reproducibility].

@@

## Removing profit from academic publishing
Taking out the profiteering from publishers will take a more concerted approach. But the reality is that we have only ourselves to blame. The publishers are able to prey on us, as biological scientists, because we are used to getting all of the "frills" associated with their publishing model. This includes the designer layout, custom websites and editorial management systems.

But the reality is that we really don't need any of these frills, and if we cared more about our science and less about the prestige that publishers have worked so hard to con us into thinking that their products have higher value, we'd all be better off. Mathematicians and physicists are way ahead of us. Given that they've shown the way, it's simply up to us to [follow their models](#overlay), and embrace the change in culture needed.

>>>>

The need for transparency in science stems from the fact that most societies are making the majority of their decisions based on evidence coming through the guidance of science. If scientific evidence is at the heart of decision making, then the collection of this evidence must be transparent to those who make resulting policies, together with those who challenge them in a democracy. Moreover, as public funded scientists we should be setting a community standard of transparency for the rest of society to follow. Publishing of science is currently in transition to address the [wicked problem](#wicked), and it is vital that the scientific community leads the way forwards, and that we are not led by for-profit publishers. One of the ways to achieve transparency is through [preregistration](#commitment) of your research project to avoid [confirmation bias](#PruittData), which is, in part, a product of commercial publishers and the [metrics they promote](#impactfactor). In order to make this effective, we need the [gatekeepers](#gatekeepers) of our journals to support the [preregistration](#commitment) of research hypotheses and methods. Right now, journals should be openly advocating and encouraging preregistration with a plan to transition their journal in future to a system embracing rigour, reproducibility, and transparency [RRT @valdez2020improving]. However, many editors are resisting this move as they feel that there is no support from the community. This may well be the case, but inequalities in science, and particularly in publishing, mean that editors can either be instruments of change or at the heart of inequality in publishing (see [Part IV](#editors) and Figure \@ref(fig:publishing-triangle)). Either our editors will lead us toward transparency, or we as a community simply need to demand that they change their practices. Currently, editors are responding to calls for transparency by making small steps [for example asking for open coding: @powers2019open], rather than adopting transparency wholesale through the [badge system](#transparent) set up by Kidwell et al. [-@kidwell2016badges].

@@

### Accepting the alternative hypothesis
At the outset of our scientific research we pose a hypothesis with the expectation that we will be able to accept or reject our null hypothesis. We often think of rejecting the null hypothesis as the only result that we are interested in, but if we only ever reported these results we would not be responsible in moving our field forwards. That is, in a world where we only report significant results (i.e. reject the null hypothesis) we would necessarily keep repeating experiments where the null hypothesis is accepted, because there would never be the evidence that the hypothesis had been previously tested in the literature. This is actually practised by the majority of scientific journals who won't consider a null result, and results in '[Publication Bias](#badscience)'. It's easy to see why this is a bad policy, but it is the prevailing culture in science publishing.

If journals only publish manuscripts that reject the null hypothesis [*cf* @franco2014publication], researchers are more likely to mine their data for positive results (P hacking), or re-write their hypothesis in order to reject the null (HARKing) [@measey2021how]. Deceptive practices such as p hacking, HARKing and [salami-slicing](#salami) are not in the interests of any journals, or the scientific project in general [@ioannidis2005why; @nissen2016publication; @forstmeier2017detecting; @measey2021how].

@@

### Novel research
One criterion for many journals is that the research should be novel. This is increasingly practised by journal editors as you move up the [Impact Factor](#impactfactor) levels. Novelty sells (just think of the meaning of “new” in newspaper), and that’s the basis for selling novel stories from higher Impact Factor journals. The perils of testing increasingly unlikely hypotheses and how this inflates Type II errors as well as increasing the proportion of Type I errors are widely acknowledged [@forstmeier2017detecting; @measey2021how]. Novelty also stifles repeatability [@cohen2017how]. If we can never repeat studies in science, then a fundamental tenet of the methodology is repressed. Reproducibility in science has received a lot of attention recently, as attempts to reproduce the results of highly cited research have failed [@opensciencecollaboration2015estimating; @begley2012raise]. This has been followed by general outrage among scientists that things should change [@anderson2007normative; @munafo2017manifesto], including a majority of those in biological sciences [@baker20161500]. The irony that these reports and requests are published in exactly the journals that will refuse to publish research that seeks to repeat work (is not novel) is clearly lost on the editors. However, more nuanced views are also coming forwards to actively introduce variable conditions and sampling of biological variation into the study design to more fully represent the nature of biological variation making studies more likely to be replicated [@voelkl2020reproducibility].

@@

## Removing profit from academic publishing {#openness}
Taking out the profiteering from publishers will take a more concerted approach, especially now that they are attempting to capture our entire work-flow [@brembs2021replacing]. But the reality is that we have only ourselves to blame. The publishers are able to prey on us, as biological scientists, because we are used to getting all of the "frills" associated with their publishing model. This includes the designer layout, custom websites, editorial management systems and now increasingly the use of free tools like Mendeley, Overleaf, Peerwith and Authorea. Indeed, these and other tools can be used as spyware to capture data from individual academics and sell it [see @brembs2021replacing]. Avoid using any publisher owned software or repository, and use only [Open Source Tools](https://101innovations.wordpress.com/) [@kramer2016innovations]. An excellent way to learn and implement the use of these tools is to form a [Open Science Community](osf.io/vz2sy/) [@armeni2021wide-scale] in your institution. There you can learn more from your peers about which tools are best used in your area and with your institutional resources, and help spread the word of the need to move toward Open Science among your colleagues.

The reality is that we really don't need any of the frills or tools that publishers offer. An entire work-flow using Open Source Tools is available, and it is up to us to make this convenient for our own use, and not fall into the publishers plan to [capture academia](http://john.measey.com/Blog/2021/09/27/Academic-capture). If we cared more about our science and less about the prestige that publishers have worked so hard to con us into thinking that their products have higher value, we'd all be better off. Mathematicians and physicists are way ahead of us. It's up to us to change our culture in order to embrace openness and transparency.

### 2021-12-15 1.05-impactfactor.Rmd
Add files via upload
# with lots of changes following copy edits and proofs

It’s a relatively simple calculation as seen in equation \@ref(eq:Impact-Factor):

\begin{equation*} 
\frac{\mbox{The sum of all  citations  in  journal X for  year Y }}{(\mbox{No. pubs in journal X for year Y-1}) + (\mbox{No. pubs in journal X for year Y-2})}
(\#eq:Impact-Factor)
\end{equation*}

Because all citations for year Y are needed before the IF can be calculated for each journal, IF for the preceding two years is typically not released until June of Y+1.

Impact Factors are published by a number of different literature databases. For example, for the impact factor calculated by the [Web of Science](https://www.webofknowledge.com/), if your journal is not even listed in the Web of Science then it will not have any Impact Factor. The Web of Science is continually policing the quality of the journals, and this means from time to time journals are excluded. This tends to happen at the lower end of the Impact Factor scale. But recently it happened to some very well known journals and there was a big stink. You can read more about this [here](https://retractionwatch.com/2020/07/13/stunned-very-confused-two-more-journals-push-back-against-impact-factor-suppression/).

Note that there are potential conflicts of interest with publishing impact factors. For example, a publisher, Elsevier, owns [Scopus](www.scopus.com) and can decide whether or not a publication can get an Impact Factor. Given two equivalent journals, do you think is it more likely that journals published by Elsevier will be included or excluded from Scopus? Organisations that are used by many of our employers in their means of evaluating our effectiveness are owned by private companies. This is certainly cause for concern. There is a group of people who are trying to replace Impact Factor with a group of other metrics, so perhaps by the time you read this Impact Factor will no longer be relevant. If you're interested read more [here](https://www.nature.com/articles/d41586-019-01643-3).

Do remember that there are a number of other citation indices including Dimensions and Google Scholar. Google Scholar doesn't currently calculate the Impact Factor for all journals, only the top 100.

@@

@@

When IF was originally devised by Eugene Garfield in 1955, it wasn’t supposed to govern the lives of academics, it was simply intended to be a way of deciding which journals to include in the Science Citation Index [@garfield1999journal]. It then became useful for librarians to help them decide which journals to keep and which to ditch under ever constrained budgets (caused by publishers' ever increasing prices). But along the way, this very simple index is now considered by many people to be a measure of quality, prestige and even academic success [@garfield1999journal]. Many people have highlighted [how wrong these beliefs are](#money), but the growing trouble is that not only have many academics been misled, but so have administrators responsible for [hiring and promotions](#DORA).

In a paper by McKiernan et al. [-@mckiernan2019use], they found that IF features in the guidelines of many university panels responsible for the fate of academics jobs and therefore lives. Worryingly, many of these institutions don’t actually talk about what IF measures. Instead they equate it with values and qualities that it certainly does not represent. Thus, you may find that your career is influenced by a simple metric that almost all who use it don’t actually understand. The undue influence on lives of scientists that IF has led directly to the San Francisco Declaration on Research Assessment (known as [DORA](#DORA)) to which many institutions and publishers have signed up. You should read this very simple declaration and find out whether your institution is a signatory. If they are, then remember to hold them to the DORA principles during any assessment that you undergo.

The Impact Factor now dominates many aspects of life for Early Career Researchers, where the pay-off for a high impact publication might make the difference between having a job or leaving science altogether. The pressure is so high that it leads to [misconduct and fraud](#pruittdata).

## 5 Year Impact Factor
Many journals report their 5 Year Impact Factor in addition the the standard 2 year timeframe (as seen in equation \@ref(eq:Impact-Factor)). This is because many disciplines, such as biological sciences, don't have maximum impact of articles within the first 2 years of publication as do subjects like medicine and physics. Although the 5 Year Impact Factor might be much more appropriate, most people ignore this metric. Although this metric might be more appropriate, because of the bad ways in which people have used impact factor, it is probably better to [push back](#pushback) against this metric as you should others.

## What can you do if you publish a journal with high IF?
Very high ranking journals for Impact Factor are publications like Cell, Nature, Science and PNAS. This is because these publications are read by a very great number of people, and so are widely cited. Articles that get published in them receive a lot of attention from the press and media. This results in the prestige that a hiring institution might be looking for. If your academics publish in this journal, your institution may well receive lots of positive publicity. In some countries, notably China, there may be a cash incentive towards publishing in a journal with a high IF [@quan2017publish].

One frightening trend that we are seeing in biological sciences is that the higher the Impact Factor the more the journal will charge you to publish in it. At the time of writing in November 2020 _Nature_ has just announced that they will charge $9500 to publish in their highest-ranked journal (see [part 2](#openaccess)). This is more money than it cost to publish in any other journal at the moment, and will be greater than the cost of many research projects, or even salaries of Early Career Researchers in some parts of the world.

@@

Some countries reward their academics if they publish in high ranking journals. This can result in a salary bonus. It may also help with promotion, getting tenure or even just getting an interview for a job. If you're going to publish and you want a career in academia then you need to be aware of Impact Factors and what they mean to different stakeholders.

@@

## Editors try to increase IF {#increaseIF}
It's important to remember that editors care about Impact Factor [see @ioannidis2019user]. There are several reasons for this. Firstly, the impact factor of the journal can be used (by the publisher or society) as a simple measure of how well the editor is doing. Secondly, the higher the impact factor of the journal, the better quality of submissions of manuscripts to publish. Being the editor of a journal with a low (or no) impact factor can result in a lot of mediocre manuscripts being submitted. Editors can only choose their content from what is submitted. Poor manuscripts take up much more time than good ones: more rounds of review, more disagreements among reviewers and more time spent making editorial decisions. Thus, by increasing the IF of the journal that you edit, you are likely to increase both the number of submissions (allowing you to reject poorer ones) and retain better ones.

All this means that if editors believe that your paper will not garner the same or more citations in two years as the current Impact Factor of their journal, they may [desk reject](#deskreject). This is just one of the ways in which editors are known to manipulate impact factors for their journals. Established ways [@metze2010bureaucrats; @martin2016editors] of editors increasing impact factor for their journals include:

1. Ask authors to cite publications from your journal published within the last 2 years. 
1. Ask reviewers to suggest publications from your journal published within the last two years to authors on which their review is conducted.
1. Encourage the submission of papers from laboratories with high output and citation rates.
1. Reject papers that are likely to have no citations. This effectively reduces the size of the denominator in the above equation.
1. Publishing issues in January means they have a maximum period of the year to get cited. This is now being inflated to having issues published online well ahead of the January date all the time gathering citations. 
1. Encourage review articles which themselves garner more citations.
1. Editorials that cite every paper in the journal. This tactic is frequently used in special issues.

@@

## Push back against IF {#pushback}
One very simple way that you can push back against industry calculated IF is to calculate IF scores for your own papers, and show how they relate to the IF of the journal that you publish in. In this way, you are simply comparing your actual citations in the 2 years after your paper is published with the mean for the journal. There is an even chance that you generally get more citations than the mean for the journal, and you can convincingly show that your citations are consistently higher than the journal IF. For this to be true, you might need to help your work get cited, and [that’s the subject of another section](#citations).

Just like any metric, Impact Factor is liable for abuse. You need to be aware of how IF is used and abused by many people in the academic community. You also need to be aware of what the rewards are for these individuals. Our problem with Impact Factor is not really the way in which it is manipulated by individuals to achieve their own ends. Instead, we should be worried about the way in which it leads the scientific community towards [bad science](#badscience) and [dishonesty](#pruittdata). Moreover, that this happens not to our benefit, but for the benefit of the publishing industry who use this metric (that they own and police themselves) to extort money from taxpayers earmarked for research into their own pockets (more on this in [Part IV](#part4)).

>>>>

It's a relatively simple calculation as seen in equation \@ref(eq:Impact-Factor):

```{=tex}
\begin{equation*} 
\frac{\mbox{The sum of all  citations  in  journal X for  year Y }}{(\mbox{No. pubs in journal X for year Y-1}) + (\mbox{No. pubs in journal X for year Y-2})}
(\#eq:Impact-Factor)
\end{equation*}
```
Because all citations for year Y are needed before the IF can be calculated for each journal, IF for the preceding two years is typically not released until June of Y+1.

Impact Factors are published by a number of different literature databases. For example, for the impact factor calculated by the [Web of Science](https://www.webofknowledge.com/), if your journal is not even listed in the Web of Science then it will not have any Impact Factor. The Web of Science is continually policing the quality of the journals, and this means from time to time journals are excluded. This tends to happen at the lower end of the Impact Factor scale. But recently it happened to some very well known journals and there was a big stink [@pinto2021are]. You can read more about this [here](https://retractionwatch.com/2020/07/13/stunned-very-confused-two-more-journals-push-back-against-impact-factor-suppression/).

Note that there are conflicts of interest with publishing Impact Factors. For example, a publisher, Elsevier, owns [Scopus](www.scopus.com) and can decide whether or not a publication can get an Impact Factor. Similarly, the new scholarly database, [Dimensions](https://www.dimensions.ai/), is owned by Holtzbrinck who also own Springer Nature. This is really only the tip of the iceberg when it comes to each publishing company's intention to capture the entire academic cycle [see @brembs2021replacing]. Hence, given two equivalent journals, do you think is it more likely that journals published by Elsevier will be included or excluded from Scopus? Databases that are used by many of our employers in their means of evaluating our effectiveness are owned by for-profit companies. This is certainly cause for concern. There is a group of people who are trying to replace Impact Factor with a group of other metrics, so perhaps by the time you read this Impact Factor will no longer be relevant. If you're interested read more [here](https://www.nature.com/articles/d41586-019-01643-3).

Note that Google Scholar doesn't currently calculate the Impact Factor for all journals, only the top 100.

@@

When IF was originally devised by Eugene Garfield in 1955, it wasn't supposed to govern the lives of academics, it was simply intended to be a way of deciding which journals to include in the Science Citation Index [@garfield1999journal]. It then became useful for librarians to help them decide which journals to keep and which to ditch under ever constrained budgets (caused by publishers' ever increasing prices). But along the way, this very simple index is now considered by many people to be a measure of quality, prestige and even academic success [@garfield1999journal]. Many people have highlighted [how wrong these beliefs are](#money), but the growing trouble is that not only have many academics been misled, but so have administrators responsible for [hiring and promotions](#DORA).

@@

In a paper by McKiernan et al. [-@mckiernan2019use], they found that IF features in the guidelines of many university panels responsible for the fate of academics jobs and therefore lives. Worryingly, many of these institutions don't actually talk about what IF measures. Instead they equate it with values and qualities that it certainly does not represent. Thus, you may find that your career is influenced by a simple metric that almost all who use it don't actually understand. The undue influence on lives of scientists that IF has led directly to the San Francisco Declaration on Research Assessment (known as [DORA](#DORA)) to which many institutions and publishers have signed up. You should read this very simple declaration and find out whether your institution is a signatory. If they are, then remember to hold them to the DORA principles during any assessment that you undergo.

> Goodheart's Law: "When a measure becomes a target, it ceases to be a good measure.”
>
> ---Marilyn Strathern [-@strathern1997improving]

Charles Goodheart originally came up with a term that specifically related to economics, but this was generalised beyond economics by Marilyn Strathern [-@strathern1997improving]. It is hard to emphasise just how far publishing metrics have now come to embody Goodheart's Law [@fire2019over-optimization]. The Impact Factor now dominates many aspects of life for Early Career Researchers, where the pay-off for a high impact publication might make the difference between having a job or leaving science altogether. The pressure is so high that it leads to [bad behaviour](#badscience), [misconduct and fraud](#pruittdata).

## Five year Impact Factor {#fiveyear}

Many journals report their 'Five Year Impact Factor' in addition the the standard two year timeframe (as seen in equation \@ref(eq:Impact-Factor)). This is because many disciplines, such as biological sciences, don't have maximum impact of articles within the first two years of publication as do subjects like medicine and physics. Papers with immediate impact can be equated with sensational breaking news stories that instantly grab headlines, making money for those news outlets. These 'hot papers', 'jump-decay' or 'fast breaking papers' (i.e. left skewed distribution) will inflate the IF of the journal, which, if sustained, will earn the publisher more [money](#money). Most papers in biological sciences have a 'delayed peak' (i.e. low frequency and long tails), consistently gathering small numbers of citations over long periods of time. Papers that suddenly become popular after a long time with no or very few citations are termed 'sleeping beauties' [i.e. right skewed, see @bornmann2012anna]. The last in the group of four citation types includes those that have a constant, usually low, number of citations over time. Together, each  of these four patterns can be described by three variables that describe the citation behaviour of all papers: fitness, immediacy and longevity [see @fortunato2018science]. Using fitness, immediacy and longevity of citations to papers within a field can help to [normalise citation metrics](#normal) between different areas of science.

Although the five year Impact Factor might be much more appropriate, most people ignore this metric. While this metric might be more appropriate, because of the bad ways in which people have used impact factor, it is probably better to [push back](#pushback) against this metric as you should against others.

## What can you do if you publish in a journal with high IF?
Very high ranking journals for Impact Factor are publications like _Cell_, _Nature_ and _Science_. This is because these publications are read by a very great number of people, and so are widely cited. Articles that get published in them receive a lot of attention from the press and media. This results in the prestige that a hiring institution might be looking for. If your academics publish in this journal, your institution may well receive lots of positive publicity. In some countries, notably China, there may be a cash incentive towards publishing in a journal with a high IF [@quan2017publish].

These highest ranking journals also demonstrate that IF is not linked to quality. Increasing numbers of disciplines are seen to be rejecting the reliability of many papers published in highly prestigious journals [@brembs2018prestigious]. Moreover, there are higher levels of fraud reported in higher ranking journals [see part IV](#fabrication). Only ~25% of articles published in these journals contribute to their very high impact factors [@editorial2005not-so-deep]. Taken together, we should question the worth of these highest ranking journals to the scientific project.

One frightening trend in biological sciences is that the higher the Impact Factor the more the journal will charge you to publish in it [@gray2020sorry]. At the time of writing in November 2020 _Nature_ has just announced that they will charge USD 9500 to publish OA in their highest-ranked journal (see [Part II](#openaccess)). These sums are an order of magnitude greater than journals like these (with rejection rates >90%) should reasonably charge [@grossmann2021current]. This is more money than it cost to publish in any other journal at the moment, and will be greater than the cost of many research projects, or even salaries of Early Career Researchers in some parts of the world.

@@

Some countries reward their academics if they publish in high ranking journals. This can result in a salary bonus [@quan2017publish]. It may also help with promotion, getting tenure or even just getting an interview for a job [@schimanski2018evaluation; @mckiernan2019use]. If you're going to publish and you want a career in academia then you need to be aware of Impact Factors and what they mean to different stakeholders.

@@

## Editors try to increase IF {#increaseIF}
It's important to remember that editors care about Impact Factor [see @ioannidis2019user]. There are several reasons for this. Firstly, the impact factor of the journal can be used (by the publisher or society) as a simple measure of how well the editor is doing. Secondly, the higher the impact factor of the journal, the number of submissions of manuscripts increases so that the editor can select ones they perceive to be of higher 'quality'. Being the editor of a journal with a low (or no) impact factor can result in receiving fewer, more mediocre manuscripts. Editors can only choose their content from what is submitted. Poor manuscripts take up much more time than good ones: more rounds of review, more disagreements among reviewers and more time spent making editorial decisions. Thus, by increasing the IF of the journal that you edit, you are likely to increase both the number of submissions (allowing you to reject poorer ones) and retain better ones.

All this means that if editors believe that your paper will not garner the same or more citations in two years as the current Impact Factor of their journal, they may [desk reject](#deskreject) your submission. This is just one of the ways in which editors are known to manipulate impact factors for their journals. Established ways [@metze2010bureaucrats; @martin2016editors] of editors increasing impact factor for their journals include:

1. Ask authors to cite publications from their journal published within the last 2 years. 
1. Ask reviewers to suggest publications from their journal published within the last two years to authors on which their review is conducted.
1. Encourage the submission of papers from laboratories with high output and citation rates.
1. Reject papers that are likely to have no citations. This effectively reduces the size of the denominator in the above equation.
1. Publishing issues in January means they have a maximum period of the year to get cited. This is now being inflated to having issues published online well ahead of the January date all the time gathering citations. 
1. Encourage review articles which themselves garner more citations.
1. Editorials that cite every paper in the journal. This tactic is frequently used in special issues.

Evidence of this unscrupulous wrangling of Impact Factor within the Biological Sciences has been found when looking at the differentiation of [inflation rates between journals](#IFinflation). Wilson [-@wilson2007journal] found that [for-profit](#money) journals managed to increase their IF faster than society journals within the same period.

@@

## Push back against IF {#pushback}
Just like any metric, Impact Factor is liable for abuse. You need to be aware of how IF is used and abused by many people in the academic community. You also need to be aware of what the rewards are for these individuals. Our problem with Impact Factor is not really the way in which it is manipulated by individuals to achieve their own ends. Instead, we should be worried about the way in which it leads the scientific community towards [bad science](#badscience) and [dishonesty](#pruittdata). People who have benefited from using IF to measure their careers are likely to object if their institutions abandon it [e.g. @chawla2021scientists], even though they are already signatories of [DORA](#DORA). Retaining IF benefits these senior academics, their closed practices and the publishing industry who use this metric (that they own and police themselves) to direct money from taxpayers earmarked for research into their own pockets (more on this in [Part IV](#part4)).

If you must calculate IF, one very simple way that you can push back against industry calculated IF is to calculate IF scores for your own papers, and show how they relate to the IF of the journal that you publish in. In this way, you are simply comparing your actual citations in the years (2 or more as appropriate) after your paper is published with the mean for the journal. There is an even chance that you generally get more citations than the mean for the journal, and you can convincingly show that your citations are consistently higher than the journal IF. For this to be true, you might need to help your work get cited, and [that's the subject of another section](#citations).
